\chapter{Implementation}
In this chapter, we describe how we went about the implementation of the
debugger and reason about the design choices we made. Also, we describe which
parts of the virtual machine were modified or added for the benefit of the
debugger or, in general, for the greater good.

\section{T86 ISA extensions}\label{section:parser}
In chapter~\ref{section:t86-vm}, we showed how to build a program for the T86
VM. It was necessary to use the builder the T86 VM provides. We remedied this
with a custom ELF-like format, which we provide a parser for. This format
semantically follows the ELF format and allows the creation of a T86
executable. Contrary to the ELF, it is a text format to allow ease of use. In
chapter \ref{section:t86-vm}, we showed how to build a program for the T86 VM
with the builder interface. To allow the usage of other programming languages,
we have created an ELF-like format for the T86 executables. The format is a
text one, making it easy to use. An example of a program in said format is
shown in figure \ref{fig:t86-program}. It is very similar to the assembly we
have shown in previous sections. Thanks to this, students can implement their
compiler in any programming language they want and emit the T86 program in this
format as a text file. An unfortunate side effect is that we can no longer use
the \texttt{DBG} instruction. However, the debugger we will later present will
be much more powerful than the \texttt{DBG} instruction.

As can be apparent from the example, we also use sections. The \texttt{.text}
section is the only mandatory one. It contains the instructions that will be
executed. Another one is the \texttt{.data} section. Here, either raw numbers
or strings can be written. The contents of this section are then loaded by the
VM and stored into the memory, beginning at memory cell 0 and upwards. There
are also debug sections, which we will present when discussing debugging
information.

\begin{figure}
    \begin{lstlisting}
.data
"Hello, World!\n"

.text
0   MOV [BP - 1], 0
1   JMP 8
2   MOV R0, [BP - 1]
3   MOV R1, [R0]
4   PUTCHAR R1
5   MOV R0, [BP - 1]
6   ADD R0, 1
7   MOV [BP - 1], R0
8   MOV R0, [BP - 1]
9   CMP R0, 13
10  JLE 2
11  HALT
    \end{lstlisting}
    \caption{Example of an T86 program which prints "Hello, World!".}
    \label{fig:t86-program}
\end{figure}
We will also add two new instructions. First is the \texttt{PUTNUM}
instruction, which prints the numerical value in the register and a newline.
This is intended as a very primitive debug instruction and to ease the
automated testing of the compiler. The only other way of output was to print a
char which was represented by the ASCII value. With this instruction, students
can bootstrap the basic implementation of their compiler more easily.

Another one is the \texttt{BKPT} instruction. This instruction is similar to
the \texttt{INT3} instruction from x86-64 or the \texttt{BKPT} instruction from
ARM. It is a software breakpoint. Control will be passed to the connected
debugger when the instruction is hit. This is thoroughly described in the next
section. 

\section{T86 debugging support}
We could bake the debugger into the virtual machine itself, which would likely
be the simplest way to implement it. However, the goal of the debugger is not
only to ease the code generation part but to be a learning point so that
students might grasp how a real debugger works\footnote{The VM followed the
same philosophy.}. Because of this, we aim to simulate the real-world debuggers
as closely as possible. The compilers may also have more targets in the future,
not just the T86 VM. If we made the debugger part of the T86 VM, we could not
use it for a possible new virtual machine. In conclusion, the virtual machine
and the debugger will be two entirely different programs and as such, two
completely different processes.

In the debugger implementation for Linux, the subject of section
\ref{section:linux-dbg}, we described how an operating system's kernel allows
the debugger's implementation via a specific API. There is no operating system
between the virtual machine and the program. Still, we will strive to make the
API similar to the ptrace API. The debugger and the VM will have to communicate
together somehow. For interprocess communication, there are several
possibilities.

Both the VM and the debugger use an abstract class representing an interface
that provides two methods, \texttt{Send} and \texttt{Receive}. The
implementation of this interface then handles the concrete way of
communication. The debugger and VM do not care about it; they merely use these
two methods. There are currently two implementations of this interface. One is
using network communication through sockets. This way, the debugger may attach
to an existing process, even on an entirely different computer. It, however,
has a disadvantage. The messages sent are often short and we need to send a lot
of them. This proved too slow, even with few messages being sent. The second
implementation is via threads. The debugger runs the VM in another thread, and
they communicate via shared queues. This is far faster and allows the debugger
to run the process by himself, making it easier to use and behave like
real-world debuggers.

The format of the communication is a text one, merely because of the ease
of use as opposed to binary format, it is also clearer to see what is
happening. The commands that the virtual machine API offers are
\begin{itemize}
    \item \texttt{PEEKREG x} - Return values of all normal registers.
    \item \texttt{POKEREG x y} - Sets the value in register \texttt{x} to
        $y$.
    \item \texttt{PEEKFLOATREG} - Return values of all float registers.
    \item \texttt{POKEFLOATREG x y} - Sets the value in float register
        \texttt{x} to $y$.
    \item \texttt{PEEKDEBUGREG} - Return value in all debug registers.
    \item \texttt{POKEDEBUGREG x y} - Sets the value in debug register
        \texttt{x} to $y$.
    \item \texttt{PEEKDATA x cnt} - Return value in memory at addresses $x$ to $x + cnt - 1$.
    \item \texttt{POKEDATA x y} - Writes a value $y$ into a memory at
        address \texttt{x}.
    \item \texttt{PEEKTEXT x cnt} - Returns instruction from $x$ to $x + cnt - 1$.
    \item \texttt{POKETEXT x INS} - Rewrite the instruction at address
        \texttt{x} with the newly supplied instruction.
    \item \texttt{CONTINUE} - Continue the execution.
    \item \texttt{TERMINATE} - End the execution.
    \item \texttt{REASON} - Get the reason why the program stopped (breakpoint,
        singlestep, halt).
    \item \texttt{SINGLESTEP} - Does native level single step.
    \item \texttt{TEXTSIZE} - Returns the size of the program.
\end{itemize}
An example of how those commands can be used for communication between the
virtual machine and the debugger is shown in figure \ref{fig:dbg-vm-seq}. The
interface is similar to basic ptrace commands. We separate the memory and
instruction writing because T86 uses Harvard architecture, whereas Linux
doesn't separate text and data address spaces\cite{ptrace}, so the two requests
were equivalent there. The API is made to be simple on purpose. Anything more
complex should be handled in the debugger itself.

The \texttt{Cpu} class is the backbone of the whole virtual machine. It is
responsible for executing the program. The \texttt{Cpu} has some light
debugging capabilities; they were mentioned in section
\ref{section:t86-debug-cap}. Unrolling must be done to display proper values in
registers and memory. This was described in section
\ref{section:superscalar-cpu}. This is already implemented in the VM by its
author, and we can use the same mechanisms for the software breakpoints and
single-stepping.

The \texttt{Cpu} class offers the \texttt{halted} function. This function
returns true if the \texttt{Cpu} executed the \texttt{HALT} instruction, which
marks the end of a program. It also has a \texttt{tick} method, which does one
tick of the CPU. This does not mean that one tick executes one instruction;
there are several pipeline stages that the instruction needs to go through. To
run the VM, one would have to write a loop that checks if the VM has halted. If
it did not, then one tick would be performed. As of now, the Cpu cannot signal
abnormal conditions (excluding halt), like breakpoint hits.

We added another manager-like class called \textit{OS}. This class will take
care of running the program via the CPU class. We also added an
\textit{interrupt} capability to the CPU. For example, when the \texttt{BKPT}
instruction is executed, CPU signals interrupt number $3$. To check if and
which interrupt happened, the CPU now provides a function for it, similar to
the \texttt{halted} function. The OS calls the \texttt{tick} method
periodically, and after every tick, it checks if a halt or interrupt occurred.
If it did, then it passes it to some handler.

When an interrupt that is caused by some debugging features happens (like the
\texttt{BKPT} instruction), the OS calls a method in the \texttt{Debug} class.
This class is also a new addition and is responsible for communication with the
debugger. It uses the text protocol we mentioned previously.

There are several features we added to the CPU to enhance debugging. One of
them is the \texttt{BKPT} instruction, which we have already talked about.
Executing this instruction causes an interrupt \texttt{3} to occur. It is also
possible to set a special flag that causes the CPU to send the interrupt
\texttt{1} after every executed instruction. 

We also added debug registers. They are a special type of registers designed
for triggering breaks on memory access. There are a total of five debug
registers, with the first four containing the memory cell addresses. The fifth
register, called the control register, contains four bits that indicate the
status of each of the first four registers. If a register is active and the
program writes to a memory cell with the same address as is stored in the
register, an interrupt \texttt{2} is generated. Furthermore, the control
register's bits from 8 to 11 reveal which register caused the interrupt. For
instance, if bit 10 is set to 1, the third register is responsible for the
interrupt and the address stored in that register is the one that was written
into.

\begin{figure}
    \centering
    \scalebox{0.8} {
    \begin{tikzpicture}
        \draw (0,0) -- (0,-17.2) (7,0) -- (7,-17.2);
        \node at (7,.3) {Debugger};
        \node at (0,.3) {Virtual machine};
        \draw[<-] (0,-1) -- node[midway,above] {Initializes connection} (7,-1);
        \draw[->] (0,-2) -- node[midway,above] {Accepts connection} (7,-2);
        \draw[<-] (0,-3) -- node[midway,above] {\texttt{PEEKTEXT 5 1}} (7,-3);
        \draw[->] (0,-4) -- node[midway,above] {Ok} (7,-4);
        \draw[<-] (0,-5) -- node[midway,above] {\texttt{POKETEXT 5 BRKPT}} (7,-5);
        \draw[->] (0,-6) -- node[midway,above] {Ok} (7,-6);
        \draw[<-] (0,-7) -- node[midway,above] {\texttt{CONTINUE}} (7,-7);
        \draw[->] (0,-8) -- node[midway,above] {Ok} (7,-8);
        \draw[->] (0,-10) -- node[midway,above] {Program stopped} (7,-10);
        \draw[<-] (0,-11) -- node[midway,above] {\texttt{REASON}} (7,-11);
        \draw[->] (0,-12) -- node[midway,above] {Reason: \texttt{BKPT} instruction} (7,-12);
        \draw[<-] (0,-13) -- node[midway,above] {\texttt{CONTINUE}} (7,-13);
        \draw[->] (0,-15) -- node[midway,above] {Program stopped} (7,-15);
        \draw[<-] (0,-16) -- node[midway,above] {\texttt{REASON}} (7,-16);
        \draw[->] (0,-17) -- node[midway,above] {Reason: \texttt{HALT} instruction} (7,-17);
    \end{tikzpicture}
    }
    \caption{A sequence diagram for the communication between the virtual machine and the debugger.}
    \label{fig:dbg-vm-seq}
\end{figure}

\section{Native Debugger}
The implementation is done in the C++ language. It uses newer standards up to
the C++20 standard. The debugger is implemented as a library. We will call this
the backend of the debugger. A command line interface was also developed,
through which the users might interact with the debugger. This will be called
the frontend of the debugger.

The implemented debugger consists of two main parts. The first one aims to
support native (instruction) level debugging. This part work without
\textbf{any} debugging information whatsoever. The second part focuses on
source-level debugging, described in the section \ref{section:source-debugger}.

The native debugger is split into two additional main layers to make it more
modular. The first layer is called a \texttt{Process}. It is an interface
representing the debuggee process. The implementation of this interface is
responsible for dealing with the concrete architecture, the API of that
architecture, and the communication with the debuggee. One implementation is
provided for the T86 VM. For instance, it has a method called \texttt{ReadText}
and \texttt{WriteText}. The internals of these methods use the
\texttt{PEEKTEXT} and \texttt{POKETEXT} API we described. Outside of this
class, the communication API is never used. If, in the future, another virtual
machine is made, for whichever architecture, it is only needed to implement
this interface. The rest of the debugger can be used as-is.

Another layer is the \texttt{Native} class, which implements the complicated
logic behind a debugger, like setting a breakpoint, handling single-step, and
so forth. It is the primary bread and butter of the native part of the
debugger. Most algorithms are similar to the Linux debugger implementation
presented in section \ref{section:linux-dbg}. For illustration, in figure
\ref{t86dbg:breakpoint} we show a snippet of code used to create a breakpoint.
It first reads the text at the address where we want to set the breakpoint. The
breakpoint opcode then rewrites this text, and the backup of the text is
stored.

When we arrive at the breakpoint and want to continue further, we need to unset
the breakpoint, i.e., replace the breakpoint opcode in the T86 program with the
backup we saved, do a native-level single step, and write the breakpoint back.

Since breakpoints change the underlying code of the debuggee, we need to be
careful when presenting information to the user. If we printed the text that we
get from the debuggee, it might contain the \texttt{BKPT} instructions we set
earlier. We need to mix it with the backup code stored in breakpoints to show
the assembly of the program correctly.

The Native class uses a \texttt{DebugEvent} structure which indicates what
caused the VM to stop. It is implemented as a \texttt{variant} of multiple
structures, for instance, the \texttt{BreakpointHit} or the
\texttt{WatchpointTrigger} structure. It is a variant because the watchpoint
also needs to convey information about an address that caused the break, as do
breakpoints. It could also signal if the break was caused by reading or writing
to the memory cell, although for now, the T86 VM only interrupts on writing.

\begin{figure}
    \begin{minted}{c++}
SoftwareBreakpoint CreateSoftwareBreakpoint(uint64_t address) {
    auto opcode = GetSoftwareBreakpointOpcode();
    // Read the text at the breakpoint address
    auto backup = process->ReadText(address, 1).at(0);
    // Rewrite it with the breakpoint opcode
    std::vector<std::string> data = {std::string(opcode)};
    process->WriteText(address, data);
    // Check that it was truly written
    auto new_opcode = process->ReadText(address, 1).at(0);
    if (new_opcode != opcode) {
        Error(...);
    }
    // Create a breakpoint object which keeps the text backup
    return SoftwareBreakpoint{backup, true};
}
    \end{minted}
    \caption{Debugger code in \texttt{Native} class to enable a breakpoint.}
    \label{t86dbg:breakpoint}
\end{figure}

The native debugger has the following features:
\begin{itemize}
    \item Breakpoints - Can set, unset, enable and disable software breakpoints.
    \item Watchpoints - Can set and unset watchpoints on memory writes.
    \item Single stepping - Can do native level step into, which executes
        current instruction, out, which runs the program until it leaves
        current function and over, which treats function calls as a single
        instruction.
    \item Text manipulation - Can read and write into the debuggee text area,
        effectively allows to overwrite the running code.
    \item Data manipulation - Can read and write into the program memory area.
    \item Register manipulation - Can manipulate with normal, float and debug registers.
\end{itemize}

\section{Source debugger}\label{section:source-debugger}
With the solid foundation represented by the native part of the debugger, we
can extend it by providing some form of source-level debugging. For this part,
we need to remember that the debugger will only be used by students. As such,
we ought to have gentler debugging information than DWARF, but we certainly can
take inspiration from it.

As we previously mentioned, the executable with T86 code is separated into
sections. The \texttt{.text} and \texttt{.data} sections are for the VM. We
will introduce new sections where debugging information will be stored. All
those sections will have \verb|.debug_| prefix. The simplest new section is
\verb|.debug_source|, which should contain the original source code which was
compiled into this executable. This later allows us, with the combination of
other information, to display the source code lines.

The main philosophy of the source-level debugger is to allow an arbitrary
amount of debug information. For instance, the user can generate information
about one function only, and for that function, source debugging capabilities
will work, but not for any other. This means that users can generate debugging
information incrementally.

The debugger is, of course, not only limited to TinyC language. Any imperative
language that can be encoded with the following debugging information is
suitable to be debugged at the source level. We show an example of this in a
provided test case for the debugger. It can be found as an attachment to the
thesis on path \todo{doplnit cestu}.

\subsection{Line information}
Line information is one of the simplest to generate. Since it will probably be
the most generated information by the students, we choose the following simple
format for it: \texttt{<line>:<address>}. The section \verb|.debug_line| is
made up of these lines. The format is self-explanatory. It says that line
\texttt{x} corresponds to address \texttt{y}. This is very easy to generate.

With this information, we are able to do source-level breakpoints. If the
source code is also provided, we can show the user on which line is the
debugged program currently paused. It is not necessary to specify every line in
the program. The debugger will refuse to put a source-level breakpoint on some
line if it does not have the necessary information.

\subsection{Debugging information format}
In the line information, we provided a straightforward format. However, we will
need a more sophisticated structure to describe some advanced constructs of the
source code. Here we will draw inspiration from the Dwarf Debugging Information
(DIE). Take a look at figure \ref{fig:t86dbg-die}, which shows an example of
such debugging information. It has a tree-like structure which, in some ways,
mimics the original program. The nodes of this tree are also called debugging
information entries (DIEs). Those entries can have other entries as their
children, and each entry has a tag that is part of its name (for example, the
\verb|compilation_unit| tag). They can also have attributes that describe their
properties. As can be seen, this is very similar to the DWARF debugging
information format. Unlike DWARF, it will be a text format. This allows us to
generate the format easily and to spot mistakes quickly. We don't want the
students to debug their generated debugging information.

For instance, the tag \verb|DIE_function| represents a function. As attributes,
it has a name, beginning address, and end address. With this additional
information, we can set a breakpoint on a function name. We can also display in
which function we are located when a break happens.

It also has one direct child, a \verb|DIE_scope|. The scope entry is mainly
used for keeping track of which variables are currently active because the T86
(or any other assembly language in general) has no notion of scopes. In the
scope entry in the example, only one variable called \texttt{d} exists. Thanks
to those entries, we can list currently active variables. We, however, often
need to examine the value of a variable. To achieve this, information about the
location and the variable type is needed.

\begin{figure}
    \begin{lstlisting}
DIE_compilation_unit: {
DIE_function: {
    ATTR_name: main,
    ATTR_begin_addr: 0,
    ATTR_end_addr: 10,
    DIE_scope: {
        ATTR_begin_addr: 0
        ATTR_begin_addr: 10
        DIE_variable: {
            ATTR_name: d,
        },
    }
}
}
    \end{lstlisting}
    \caption{Debugging function information for the T86 debugger.}
    \label{fig:t86dbg-die}
\end{figure}

The type information is encoded as a standalone DIE. Currently, three type
entries are present, one for primitive types (int, double, or char), one for
pointers, and one for structured types (\texttt{struct} or \texttt{class} in
C++). Other types can be easily added in the future. The types are saved as
separate entries, and as such, we need some way to link them together with the
variables. We will use the \verb|ATTR_id| attribute to achieve this. This
attribute should be unique for every entry. This role is similar to the
\texttt{id} attribute of HTML elements~\cite{html4}. The variables themselves
have the \verb|ATTR_type| attribute, which will have an id of the type as its
value. An example of a pointer type that points to an int type is in figure
\ref{fig:t86dbg-types}. If we had a variable that is a pointer to int, it would
need to have the \verb|ATTR_type: 1| attribute because the id of a pointer type
to integer is one.

The primitive types need to have their size. For T86, this is the number of
memory cells it occupies, which will almost always be one since one memory cell
is 64 bits. It also has a name for its primitive type. Currently, three
are supported:
\begin{itemize}
    \item \texttt{int} - A signed integer.
    \item \texttt{float} - A floating point number.
    \item \texttt{char} - A number representing an ASCII character.
\end{itemize}

Structured types are more complicated. They need to have a list of members
which are stored in the structure. For each member, an offset from the
beginning of the structure must also be specified. It also must provide a size
because the compiler might align it, and it may be larger than the sum of the
size of its members.

\begin{figure}
    \begin{lstlisting}
DIE_primitive_type: {
    ATTR_name: int,
    ATTR_id: 0,
    ATTR_size: 1,
},
DIE_pointer_type: {
    ATTR_type: 1,
    ATTR_id: 1,
    ATTR_size: 1
},
    \end{lstlisting}
    \caption{Debugging type information for the T86 debugger.}
    \label{fig:t86dbg-types}
\end{figure}

With this information, we can show the type of the variable. Nevertheless, the
most valuable thing is its value. Variables are either stored in memory,
registered, or optimized out completely. We will follow DWARF's footsteps and
provide a language for a virtual machine that one can interpret to gain
information about the variable location.

Several examples can be seen in figure \ref{fig:t86dbg-vm}. Same as DWARF, it
is a stack-based machine. After computation is done, the top value on the stack
is considered to be the location of the variable. The instruction \texttt{PUSH}
pushes either offset or register onto the stack. In the first example, the
final location is the register \texttt{R0}. In the second example, the location
is a memory offset equal to the value in register \texttt{BP} minus two. The
third example has the same meaning, but a shorthand was provided since the
variables will most often be stored at some offset from the register
\texttt{BP}. There is also a dereference instruction, which can be used to
inspect to which location a pointer points.

\begin{figure}
    \begin{lstlisting}
- [PUSH R0]
- [PUSH BP; PUSH -2; ADD]
- [BASE_REG_OFFSET -2]
    \end{lstlisting}
    \caption{Debugging type information for the T86 debugger.}
    \label{fig:t86dbg-vm}
\end{figure}

This information is stored in a variable attribute called
\texttt{ATTR\_location}. If all this information is provided, we know where the
variable is stored and may look up its value. Together with the type
information, we might also properly interpret the value and report it to the
user.

\subsection{Expressions}
We could make a very straightforward implementation of getting variable value
by its name. It is only a matter of finding the variable DIE with the correct
name and interpreting its location and type. However, we often need to inspect
some more complicated expressions. For example, we may want to display some
struct member or a value at which pointer points.

The debugger has a built-in interpreter for such expressions. It builds an AST
from the expression and interprets it using
AST-walk~\cite{crafting-interpreters}. It leverages the \texttt{Native} class
to fetch variable values. It supports almost all C operators, excluding the
assignment operator. An example of such an expression is \texttt{foo[2]->bar + 3}.

\subsection{Frontend}
Finally, we provide a command line interface that leverages the debugger
library to make the debugger usable. It provides many commands, and its manual
can be found as an appendix to this thesis\todo{DODAT}.

The main priority of the CLI is to make the debugger easy to use. It consists
of several commands, one of them is \texttt{breakpoint set 5}, which will set a
source-level breakpoint on the fifth line of the program. It is, however, not
necessary to write the whole command. Any prefix will do, like \texttt{b s 5}.
The CLI leverages the \textit{linenoise}~\cite{linenoise} library to make the
REPL satisfying to use.

The CLI also displays various information on program stop, like why the program
stopped, on which address or line, and prints the surrounding lines of assembly
or source. The CLI can also list breakpoints and display their locations in the
disassembly or the source code. Figure \ref{fig:cli-hit} provides an example of
a breakpoint hit report. The debugger had all debugging information available
here. It can show the line in the source code where the breakpoint happened,
name the offending function, and variables in scope.

\begin{figure}
    \begin{lstlisting}
Process stopped, reason: Software breakpoint hit at line 11
function main at 7-18; active variables: a, b
      9:    int b = 6;
     10:    swap(&a, &b);
@->  11:    print(a);
     12:    print(b);
     13:}
    \end{lstlisting}
    \caption{Example of the CLI reporting a breakpoint hit.}
    \label{fig:cli-hit}
\end{figure}
